Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-01 numerical-radiation-fld.tex

In addition to the ray-tracing approach for radiation transport
described in Section \ref{sec.num.raytracing}, Enzo currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-SAMR grids).  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

@@

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(eq.~\ref{eq:species_evolution} is solved
independently of Enzo's hydrodynamics, gravity and dark-matter solvers
(sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of equations (\ref{eq:fld_radiation}) and
(\ref{eq:fld_chemistry}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

We then evolve the system (\ref{eq:fld_heating}-\ref{eq:fld_chemistry}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c +
    \Gamma\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \frac{\partial {\mathrm n}_i}{\partial t} &=& k_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved {\em analytically} for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

>>>>

In addition to the ray-tracing approach for radiation transport
described in Section~\ref{sec.num.raytracing}, \enzo\ currently includes
a field-based radiation transport solver for problems posed on uniform
(i.e.~non-AMR grids).  Detailed explanations of the model and
solution approach may be found in
\cite{NBHBROW2007}, \cite{ReynoldsHayesPaschosNorman2009} and
\cite{NRS2009}, the salient features of which are reproduced here. 
In addition, comparisons of this solver with other astrophysical
radiation transport solvers may be found in \cite{IlievEtAl2009}.

@@

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(Equation~\ref{eq:species_evolution}) is solved
independently of Enzo's hydrodynamics, gravity and dark-matter solvers
(Sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of Equations (\ref{eq:fld_radiation}) and
(\ref{eq:species_evolution}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

We then evolve the heating and chemistry system, Equations~(\ref{eq:fld_heating})
and (\ref{eq:species_evolution}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c +
    \Gamma\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \frac{\partial {\mathrm n}_i}{\partial t} &=& k_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved {\em analytically} for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

### 2013-05-01 numerical-radiation-raytracing.tex

We solve the radiative transfer equation in comoving coordinates (given by eq.~\ref{eq:rteqn}).
We can make some appropriate approximations to reduce the complexity
of this equation in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}
where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell, as we will show later, provides an accurate calculation
of the radiation field.

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (ph s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

The rays are traced through the grid in a typical fashion
\citep[e.g.][]{Abel99_RT}, in which we calculate the next cell
boundary crossing.  The ray segment length crossing the cell is
%
\begin{equation}
  \label{eqn:trace}
  dr = R_0 - \min_{i=1 \rightarrow 3} \left[(x_{\rm cell,i} - x_{\rm src,i}) /
    \hat{n}_{\rm ray,i} \right],
\end{equation}
%
where $R_0$, $\hat{n}_{\rm ray}$, $x_{\rm cell,i}$, and $x_{\rm
  src,i}$ are the initial distance traveled by the ray, normal
direction of the ray, the next cell boundary crossing in the $i$-th
dimension, and the position of the point source that emitted the ray,
respectively.  However before the ray travels across the cell, we
evaluate the ratio of the face area $A_{\rm cell}$ of the current cell
and the solid angle $\Omega_{\rm ray}$ of the ray,
%
\begin{equation}
  \label{eqn:split}
  \Phi_c = \frac{A_{\rm cell}} {\Omega_{\rm ray}} = 
  \frac{N_{\rm pix} (\Delta x)^2} {4\pi R_0^2}.
\end{equation}
%
If $\Phi_c$ is less than a pre-determined value (usually $>3$), the
ray is split into 4 child rays.  The pixel numbers of the child rays
$p^\prime$ are given by the ``nested'' scheme of HEALPix at the next
level, i.e. $p^\prime = 4 \times p + [0,1,2,3]$, where $p$ is the
original pixel number.  The child rays (1) acquire the new normal
vectors of the pixels, (2) retain the same radius of the parent ray,
and (3) gets a quarter of the photon flux of the parent ray.
Afterward the parent ray is discontinued.

A ray propagates and splits until (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.

@@

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used to update the ionization state and
energy of the cells with the solver described in \S\ref{sec.ov.chem}.

>>>>

We solve the radiative transfer equation in comoving coordinates (given by Equation~\ref{eq:rteqn}).
We can make some appropriate approximations to reduce the complexity
of this equation in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}

where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cel  are required to provide an accurate calculation
of the radiation field (as we will show later).

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (photon s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

The rays are traced through the grid in a typical fashion
\citep[e.g.][]{Abel99_RT}, in which we calculate the next cell
boundary crossing.  The ray segment length crossing the cell is
%
\begin{equation}
  \label{eqn:trace}
  dr = R_0 - \min_{i=1 \rightarrow 3} \left[(x_{\rm cell,i} - x_{\rm src,i}) /
    \hat{n}_{\rm ray,i} \right],
\end{equation}
%
where $R_0$, $\hat{n}_{\rm ray}$, $x_{\rm cell,i}$, and $x_{\rm
  src,i}$ are the initial distance traveled by the ray, normal
direction of the ray, the next cell boundary crossing in the $i$-th
dimension, and the position of the point source that emitted the ray,
respectively.  However before the ray travels across the cell, we
evaluate the ratio of the face area $A_{\rm cell}$ of the current cell
and the solid angle $\Omega_{\rm ray}$ of the ray,
%
\begin{equation}
  \label{eqn:split}
  \Phi_c = \frac{A_{\rm cell}} {\Omega_{\rm ray}} = 
  \frac{N_{\rm pix} (\Delta x)^2} {4\pi R_0^2}.
\end{equation}
%
If $\Phi_c$ is less than a pre-determined value (usually $>3$), the
ray is split into 4 child rays.  The pixel numbers of the child rays
$p^\prime$ are given by the ``nested'' scheme of HEALPix at the next
level, i.e. $p^\prime = 4 \times p + [0,1,2,3]$, where $p$ is the
original pixel number.  The child rays (1) acquire the new normal
vectors of the pixels, (2) retain the same radius of the parent ray,
and (3) gets a quarter of the photon flux of the parent ray.
Afterward, the parent ray is discontinued.

A ray propagates and splits until at least one of the following conditions is met: (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.

@@

The radiation field is calculated by integrating Equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization and
photo-heating rates associated with a single ray ($k_{\rm ph}$ and
$\Gamma_{\rm ph}$, respectively)  are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used to update the ionization state and
energy of the cells with the solver described in Section~\ref{sec.ov.chem}.
